Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-08 article/main.tex
écrit un bout d'intro et réorganisé des trucs

%Ceux qu'on utilisait dans le projet :

\usepackage{algorithm}
\usepackage{algorithmic}

@@

@@

Today's world has become a data driven world. Having access unpredictable data is invaluable when considering security problem. For example, security keys have to be generated randomly whether we are talking about RSA keys for secure communication or a cipher key for encrypting data. For this purpose, random data can be generated by pseudo-random generators, that rely on a predetermined mathematical structure to output seemingly random data. However, by knowing the internal structure and state of the generator, it is possible to reconstruct the entire string of data.\\

The goal of our project is to study the flaws and try to cryptanalyse pseudo-random generators from the Permuted Congruential Generator (PCG) family, designed by Melissa E. O'Neill.\cite{melissapaper} Indeed, predicting a sequence from these generators is said to be "challenging"\cite{melissaweb} by their author, but this statement is not backed by any studies. Furthermore, PCG generators are very close to another well known family of generators, the Truncated Linear Congruential Generators, for which a certain amount of viable prediction algorithms are known. The combination between being highly commended (it achieved 5th place out of 29 in a recent performance test survey\cite{survey}) and a well presented website\cite{melissaweb} giving an feeling of professionalism can lead to the impression that PCG is a sufficiently secure family of generators. Given the lack of studies on the subject, there is a problem if it is used in critical pieces of software. This is why we will attempt to find flaws in PCG random generators.\\

Some attacks against PCG have been carried out by Sebastiano Vigna\cite{vignaweb}, but this didn't yield any significant results, given that he only tried to break an incomplete version in a fairly straightforward way. He worked on a very specific version of PCG. He considers the result without modulo, which is a big simplification. Moreover, he didn't study the 128 bits long generator, but only the 64 bits long\cite{vignacode}.\\

We focused on the general version of PCG, which happens to be very similar to the generator implemented by the function \texttt{random} of the C library. Both are based on a well known family of generators, the Linear Congruential Generators (LCG), but with the PCG one also using a non linear function and a rotation on the output.\\

We are looking for a possible way to reduce this generator to an LCG, or a truncated LCG, because there are known algorithms to predict them. Because we aren't studying the same exact version as Sebastiano Vigna, his findings aren't very relevant to us.\\

We decided to study the default version of the largest available generator of the 0.94 version of the PCG family, which has an internal state of 128 bits, named PCG128. We will mathematically analyse the algorithm to uncover weaknesses, and if methods for the LCG family can be translated to the PCG family.

@@

@@

\subsection{Brute-Force Attack with known $C$}
In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$ with the given formula, and predict all the following outputs of the generator. The naive way is by doing a comprehensive research of $upS_0$, and then calculating $lowS_0$ with the formula :\\
\begin{equation}
    lowS_0 = upS_0\ \mathtt{XOR}\ \mathtt{unrotate}(X_0)
\end{equation}
\texttt with {unrotate} being the inverse function of \texttt{rotate}.\\

@@

In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us an answer within $2^{128+6}$ simple operations whit the following algorithm:
\\
\begin{algorithmic}[!h]
\FOR{all possibilities of $lowC$ and $upS_0$}
    \STATE \texttt{unrotate}($X_0$)
    \STATE $lowS_0 = upS_0\ \mathtt{XOR}\ X_0$
    \STATE $lowS_1 = lowC + lowS_0 \times A \mod{2^{k/2}}$
    \FOR{all possibilities of $rot_1$}
        \STATE $\mathtt{unrotate}(X_1)$
        \STATE $upS_1 = lowS_1\ \mathtt{XOR}\ X_1$
        \STATE $C = S_1 - A \times S_0 \mod{2^{k}}$
        \STATE $S_2 = S_1 \times A + C$
        \IF {$S_2$ is coherent with  $X_2$}
            \STATE print($C$,$S_0$)
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}

@@

\subsection{Reduction to a truncated LCG model}
By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed. We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\

\begin{align}
    S_{i+1} &= A \times S_i + C \mod{2^k}\\
    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
\end{align}
We can see that $lowS_{i+1}$ is independent from $upS_i$. That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$. This relationship is used in the Brute-Force attack with known $C$, in the previous section.
We can see on (Figure \ref{pcg128goodbits})%figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
    \caption{PCG128 : Relevant bits of the outputs}
    \label{pcg128goodbits}
\end{figure}
that we can extract a lot of information by brute forcing certain well chosen bits. We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.
\newpage

\section{Prediction of truncated LCG algorithms}

\subsection{Different solutions for Truncated Linear Congruential Algorithms}

The truncated LCG are not used anymore because there exists well known solutions for predicting their data stream in polynomial time.\\

The first article we found was written by Joan Boyar\cite{Boyar1989}. The subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$ unknown and has access to a very large part of the internal state. This made it so that we couldn't use her findings, because guessing a large part of the internal state wouldn't be cost-effective. However, her study included a very broad state of the art, explaining the specifics of many algorithms, as well as describing in which circumstances they should be used.\\

The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$ and $A$ are known, $M$ being a power of two and, possibly with a significant portion of the internal state unknown. The known part is the upper bits of the state. This algorithm is coherent with our problem, this being the reason why we focused on this one in particular. However, this algorithm won't work for some particular multipliers. But this problem doesn't concern us, because the default multiplier of \texttt{PCG128} is not one of them.\\

@@

\section{Warm-Up: Prediction with Known Increment }
\subsection{Reduction to a geometric sequence and validation of our values}

@@

@@

\section{Prediction with Unknown Increment}

>>>>

\usepackage{algorithmic}

@@

\newcommand{\todo}[1]{\textcolor{red}{TODO:[#1]}}

@@

\todo{Ca, c'est du blabla...}

Pseudo-random generators (PRG) are well-studied primitives in symmetric
cryptography. A PRG is an efficient deterministic algorithm that stretch a small
random seed into a longer pseudo-random stream. To achieve cryptographic-grade
pseudo-randomness, a PRG must ensure that the pseudo-random stream is
computationally indistinguishable from a ``truly'' random sequence of bits by
efficient adversaries. Alternatively, it is possible to define pseudo-randomness
by asking that no efficient algorithm is capable of predicting the next
pseudo-random bit with non-negligible accuracy. The two definitions are in fact
equivalent.

\todo{mentioner stream ciphers}

Not all pseudo-random generators are of cryptographic strength. In some
applications, it is not necessarily necessary: to be used in Monte-Carlo
simulations or generate random choices in games, a relaxed, non-cryptographic
notion of pseudo-randomness may be sufficient. This allows for faster
algorithms. For instance, the \texttt{Math.Random} function in Google's V8
open-source JavaScript engine uses the \textsf{XorShift128} generator, which is
a Linear Feedback Shift Register tailored for high software speed. On 64-bit
processors, it produces 64 pseudo-random bits with 3 shifts and 4 XORs
instructions. The \textsf{python} standard library's \texttt{random} module uses
the Mersenne Twister. The \textsf{C} library that comes along \texttt{gcc} (the
\texttt{glibc}) uses a (poor) truncated linear congruential generator by default
to implement the \texttt{rand} function.

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' it is passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites (\todo{les
  mentionner}). The goal of designers then consists in designing the fastest
possible generator that passes the day's favorite test suite.

\todo{Mentionner Ferrenberg 92}

The scientific computing community also realized that the need for fast
\emph{parallel} random number generation could be satisfied by the use of block
ciphers in counter mode~\cite{Salmon11}. The need for speed then leads to the
use of weakened cryptographic primitives.

In most cases, it is easy to see that a non-cryptographic PRG does not meet the
cryptographic notion of pseudo-randomness, and there are few exceptions. In this
paper, we study the \textsf{PCG} family of non-cryptographic pseudo-random
generators~\cite{melissapaper,melissaweb}.

\textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
consists in applying a non-linear filtering function on top of a (fairly weak)
linear congruential generator (in a way reminiscent to the venerable filtered
LFSRs). The resulting combination is fast and passes current test suites. Its
designer claimed that distinguishing the pseudo-random stream produced would be
``challenging''. The \textsf{PCG} family contains many members, but we focus on
its strongest member, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has
a 128-bit internal state and produces 64 bits when clocked. It is the default
pseudo-random number generator in the popular \textsf{NumPy} scientific
computing package for \textsf{Python}.

The internal state of the \textsf{PCG64} generator is made of a 128-bit
``state'' and a 128-bit ``increment'', whose intended use is to provide several
pseudo-random streams with the same seed (just as the initialization vectors do
in stream ciphers). A default increment is provided in case the end-user just
want one pseudo-random stream with a single 128-bit seed.

\paragraph{Contribution.} We describe an algorithm that reconstructs the full
internal state of the strongest member of the \textsf{PCG} family. This allows
to predict the pseudo-random stream deterministically and clock the generator
backwards. The original seeds can also easily be reconstructed. The algorithm is
practical and we have executed it in practice. It follows that predicting the
output of the \textsf{PCG} should not be considered challenging anymore.

Our algorithm reconstruct the internal state using the ``guess-and-determine''
technique: some bits of the internal state are guessed ; assuming the guesses
are correct, some other information is computed ; a consistency check discards
bad guesses early on ; then candidate internal states are computed and fully
tested. The problem actually come in two distinct flavors.

When the increment is known (for instance when it is the default value), a
simplified prediction algorithm recovers the internal state from 192 bits of
pseudo-random stream. The process runs in 20 CPU minutes (on a single core of a
server processor). It guesses 36 bits of the internal state, then solves an
instance of the \textsc{Closest Vector Problem} (CVP) in a 3-dimensional
euclidean lattice. This requires about 50 arithmetic operations in total and
reveals the entire internal state if the guesses are correct.

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialized using an external source of entropy. In this case, our prediction
algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves and instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

We implemented the algorithms in \textsf{C}, then asked the designer of the PCG
family to send us ``challenge'' pseudo-random streams ; we ran our code and
emailed back the seeds used to generate the challenge streams the next day.

\paragraph{Related Work.} \todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}

\todo{mentionner Vigna ?}

% The combination between being highly commended (it achieved 5th place out of 29
% in a recent performance test survey\cite{survey}) and a well presented
% website\cite{melissaweb} giving an feeling of professionalism can lead to the
% impression that PCG is a sufficiently secure family of generators.

% Some attacks against PCG have been carried out by Sebastiano
% Vigna\cite{vignaweb}, but this didn't yield any significant results, given that
% he only tried to break an incomplete version in a fairly straightforward way. He
% worked on a very specific version of PCG. He considers the result without
% modulo, which is a big simplification. Moreover, he didn't study the 128 bits
% long generator, but only the 64 bits long\cite{vignacode}.

@@

\todo{cette section doit décrire précisément PCG.}

\begin{quote}
  In addition, one of the generators, PCG-XSL-RR(described in Section 6.3.3), is
  explicitly designed to make any attempt at state reconstruction especially
  difficult, using xor folding tominimize the amount of information about
  internal state that leaks out.
\end{quote}

\begin{quote}
  Although these properties make it extremely challenging for an agent observing
  theexternal behavior of a program to guess the state of its random number
  generator,
\end{quote}

% import numpy
% print(numpy.random.default_rng())
% ---> Generator(PCG64)

@@

\section{Naive Brute-Force Reconstruction of the Internal State}

In this section we describe the simplest possible prediction algorithm we could
think of. They serve as ``baseline'' to evaluate the more refined predictors
described below.

\todo{Dans le cas où C est connu, c'est le truc de Vigna, non ?}

\subsection{Brute-Force Attack with known $C$}

In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$
with the given formula, and predict all the following outputs of the
generator. The naive way is by doing a comprehensive research of $upS_0$, and
then calculating $lowS_0$ with the formula :
\begin{equation}
    lowS_0 = upS_0\ \mathtt{XOR}\ \mathtt{unrotate}(X_0)
\end{equation}
with \texttt {unrotate} being the inverse function of \texttt{rotate}.

@@

In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us
an answer within $2^{128+6}$ simple operations whit the following algorithm:

\begin{algorithmic}[!h]
\FOR{all possibilities of $lowC$ and $upS_0$}
    \STATE \texttt{unrotate}($X_0$)
    \STATE $lowS_0 = upS_0\ \mathtt{XOR}\ X_0$
    \STATE $lowS_1 = lowC + lowS_0 \times A \mod{2^{k/2}}$
    \FOR{all possibilities of $rot_1$}
        \STATE $\mathtt{unrotate}(X_1)$
        \STATE $upS_1 = lowS_1\ \mathtt{XOR}\ X_1$
        \STATE $C = S_1 - A \times S_0 \mod{2^{k}}$
        \STATE $S_2 = S_1 \times A + C$
        \IF {$S_2$ is coherent with  $X_2$}
            \STATE print($C$,$S_0$)
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}

@@

\section{Predicting Truncated Linear Congruential Generators}

\todo{Comme on utilise à plusieurs reprise l'idée de se ramener à un LCG
  tronqué, autant décrire la technique à base de réseaux un bon coup.}

The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$
and $A$ are known, $M$ being a power of two and, possibly with a significant
portion of the internal state unknown. The known part is the upper bits of the
state. This algorithm is coherent with our problem, this being the reason why we
focused on this one in particular. However, this algorithm won't work for some
particular multipliers. But this problem doesn't concern us, because the default
multiplier of \texttt{PCG128} is not one of them.

The first article we found was written by Joan Boyar\cite{Boyar1989}. The
subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$
unknown and has access to a very large part of the internal state. This made it
so that we couldn't use her findings, because guessing a large part of the
internal state wouldn't be cost-effective. However, her study included a very
broad state of the art, explaining the specifics of many algorithms, as well as
describing in which circumstances they should be used.

%\subsection{Different solutions for Truncated Linear Congruential Algorithms}

@@

\section{Predicting PCG with Known Increment}
\subsection{Reduction to a truncated LCG model}
By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed. We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\

\begin{align}
    S_{i+1} &= A \times S_i + C \mod{2^k}\\
    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
\end{align}
We can see that $lowS_{i+1}$ is independent from $upS_i$. That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$. This relationship is used in the Brute-Force attack with known $C$, in the previous section.
We can see on (Figure \ref{pcg128goodbits})%figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
    \caption{PCG128 : Relevant bits of the outputs}
    \label{pcg128goodbits}
\end{figure}
that we can extract a lot of information by brute forcing certain well chosen bits. We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.

\subsection{Reduction to a geometric sequence and validation of our values}

@@

[insert here] attack predicting algo description.

@@

\section{Predicting PCG with Unknown Increment}

[insert here] attack predicting algo description. Phase 1, Phase 2, Phase 3
